---
title: "Toronto Transit Commission (TTC) Streetcar Delay"
author: "Ruxin Liu"
date: "11/5/2020"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.height = 4, fig.width = 6, fig.align = "center", fig.pos = 'H')
```

```{r, echo = FALSE, message = FALSE}
delay_14 <- read.csv("ttc-streetcar-delay-data-2014.csv", header = TRUE)
delay_15 <- read.csv("ttc-streetcar-delay-data-2015.csv", header = TRUE)
delay_16 <- read.csv("ttc-streetcar-delay-data-2016.csv", header = TRUE)
delay_17 <- read.csv("ttc-streetcar-delay-data-2017.csv", header = TRUE)
delay_18 <- read.csv("ttc-streetcar-delay-data-2018.csv", header = TRUE)
delay_19 <- read.csv("ttc-streetcar-delay-data-2019.csv", header = TRUE)
delay_20 <- read.csv("ttc-streetcar-delay-data-2020.csv", header = TRUE)

colnames(delay_20)[7] <- "Min.Delay"  
colnames(delay_20)[8] <- "Min.Gap"  

delay <- rbind(delay_14, delay_15, delay_16, delay_17, delay_18, delay_19, delay_20)

library(lubridate)
library(tidyverse)
library(dplyr)
delay$Report.Date <- ymd(delay$Report.Date)
delay$year <- year(delay$Report.Date)
delay$month <- month(delay$Report.Date)

weather <- read.csv("Weather.csv")
weather <- weather[-c(13 : 84), ]
colnames(weather)[1] <- "month"
delay_final <- left_join(delay, weather, by = "month")

delay_final <- delay_final[-82703, ]

for(i in 1 : nrow(delay_final)) {
  if(delay_final$month[i] == 1) {
    delay_final$month[i] <- "Jan"
  }
  if(delay_final$month[i] == 2) {
    delay_final$month[i] <- "Feb"
  }
  if(delay_final$month[i] == 3) {
    delay_final$month[i] <- "Mar"
  }
  if(delay_final$month[i] == 4) {
    delay_final$month[i] <- "Apr"
  }
  if(delay_final$month[i] == 5) {
    delay_final$month[i] <- "May"
  }
  if(delay_final$month[i] == 6) {
    delay_final$month[i] <- "Jun"
  }
  if(delay_final$month[i] == 7) {
    delay_final$month[i] <- "Jul"
  }
  if(delay_final$month[i] == 8) {
    delay_final$month[i] <- "Aug"
  }
  if(delay_final$month[i] == 9) {
    delay_final$month[i] <- "Sep"
  }
  if(delay_final$month[i] == 10) {
    delay_final$month[i] <- "Oct"
  }
  if(delay_final$month[i] == 11) {
    delay_final$month[i] <- "Nov"
  }
  if(delay_final$month[i] == 12) {
    delay_final$month[i] <- "Dec"
  }
}

delay_final$weekend <- ifelse(delay_final$Day == "Sunday" | delay_final$Day == "Saturday", 1, 0)

route_ttc <- c(501, 503, 504, 505, 506, 508, 509, 510, 511, 512)
delay_final <- delay_final %>% 
  filter(Route %in% route_ttc)
```


# Abstract

  In the city of Toronto, it is very common to experience streetcar delays and it 
  would be helpful to understand what factors could impact the delayed time. In
  order to answer the question, a multilevel negative binomial model is fitted in 
  this study. The results found that both the delay-causing incident and the 
  delay-happening day have impact on how long the delay in minutes will be. 
  After improving this model, it will be useful for passengers to have a general
  idea about their waiting time. 


# Introduction


  When walking in the street in Toronto, Canada, it is very common to see the red
  and white streetcars driving around. These streetcars are operated by the Toronto
  Transit Commission (TTC) and there are in total 10 different streetcar routes, 
  which bring large convenience and efficiency to the transportation system in the
  city of Toronto (Wikipedia 2020). However, as many other public traffic systems 
  do, TTC streetcars also have frequent delays. The City of Toronto's Open Data 
  Portal has collected TTC streetcar delay information from the year 2014 to the 
  year 2020, which includes the location, the incident, the date, the route, the 
  length of the delay and etc (Open Data Portal 2020). 
  
  
  In this study, the major question is that what factors could potentially impact
  the delayed time. In order to answer this, a multilevel negative binomial regression 
  is fitted to the data to explore the relationships between the delayed incident, 
  the delayed day, and the length of the delay, while accounting for different streetcar
  routes. 
  
  
  
# Method

### Data Content

  The data was downloaded from The City of Toronto's Open Data Portal and was cleaned
and processed in R (all the detailed codes are in the file: Supplementary R code.Rmd).
The data was subsetted on the 10 routes that are part of the Toronto streetcar system,
which are route number 501, 503, 504, 505, 506, 508, 509, 510, 511 and 512. In 
total, there are 76886 observations. Table 1 below displays the major information 
collected in this data, where Min.Delay is the length of the delay in minutes to
the schedule for the following streetcar.


```{r, warning = FALSE, message = FALSE, echo = FALSE}
library(kableExtra)
kable(head(delay_final[c(1, 500, 1000, 9900, 13000), c(1, 2, 3, 4, 6, 7)]), 
      caption = "Data Content")
```


### Variable Selection

  In order to explore the potential factors that can impact the delayed time, 
  exploratory data analysis (EDA) is performed. From Fig.1, it is very clear that
  the distribution of the delayed time related to different delay-causing incidents
  varies a lot, especially when there is a diversion happening, the delayed time 
  is much longer compared to the other incidents. Also, lots of the outliers are 
  observed from the plot.
  

```{r, warning = FALSE, message = FALSE, echo = FALSE}
ggplot(data = delay_final, aes(x = Incident, y = Min.Delay)) + 
  geom_boxplot() +
  coord_flip() +
  ylim(0, 200) +
  xlab("Delayed Incident") +
  ylab("Delay in Minutes") +
  ggtitle("Fig.1 Relationship Between Incident & Delay Length") +
  theme(plot.title = element_text(hjust = 0.5))
```


  From Fig.2, although for all 7 days in the week, there are many outliers and 
  the 3rd quartiles are relatively at the same level, the 1st quartiles of Saturday
  and Sunday are higher compared to the other days. Therefore, an variable "weekend" 
  is created to indicate whether the day is a weekend or not to capture the difference
  in streetcar delay time between weekdays and weekends.
  
  
```{r, warning = FALSE, message = FALSE, echo = FALSE}
ggplot(data = delay_final, aes(x = Day, y = Min.Delay)) + 
  geom_boxplot() +
  coord_flip() +
  ylim(0, 50) +
  xlab("Day") +
  ylab("Delay in Minutes") +
  ggtitle("Fig.2 Relationship Between Delay Day & Delay Length") +
  theme(plot.title = element_text(hjust = 0.5))
```


### Model Selection & Validation 

  The response variable in this project is the length of delay in minutes (Min.Delay), 
  which is a continuous and numerical variable. However, the data does not have a 
  normal distribution even after transformations, and since the delayed time is 
  recorded to the nearest minute, the variable Min.Delay performs like a discrete 
  variable. Therefore, the response variable is considered as the count data (the
  number of delayed minutes) and the negative binomial regression is performed 
  instead of the linear model. The negative binomial regression includes the incident
  and weekend variables as the predictors and they are all statistically significant
  at the $\alpha$ level of 5%, which confirms with the EDA that they are important
  factors to consider. And by checking the residuals and plotting Rootograms, the 
  negative binomial model is confirmed to be the more appropriate model for this data 
  (The detailed model validation process is in the Appendix). 
  
  
  From Fig.3, it is clear that the distributions of the delayed time are quite 
  different among the 10 routes, where the red dashed line is the mean minutes of 
  delay. In order to capture this in the model, each route will have its own intercept.
  Therefore, the final model for this study is a multilevel negative binomial regression
  fitted with the *lme4* package in R, where the delayed time in minutes is the response, 
  delayed incident and weekend indicator are the predictors, with each route having
  a varying intercept (glmer.nb(Min.Delay ~ Incident + weekend + (1 | Route))).
  
  
```{r, warning = FALSE, message = FALSE, echo = FALSE}
route <- delay_final %>% 
  filter(Min.Delay < 60) %>% 
  filter(is.na(Route) == F)

ggplot(route) +
  geom_density(alpha = 0.3) +
  aes(x = Min.Delay, color = factor(Route)) + 
  facet_wrap(~ factor(Route)) + 
  geom_rug() +
  xlab("Delay in Minutes") +
  theme(legend.position="none") +
  ggtitle("Fig.3 Distribution of the delay length among different routes") +
  geom_vline(xintercept = mean(route$Min.Delay), color = "red",lty = 2) +
  theme(plot.title = element_text(hjust = 0.5))
```


# Result

  
  Based on the results of the multilevel negative binomial model, each route has 
  a slightly different intercept and the estimated average intercept is 4.34, which
  means that when the delayed incident is Diversion (the reference group) and the
  delay happens on weekdays (weekend = 0), the delayed time is expected to be 76.8
  minutes ($e^{4.34}=76.8$). From Fig.4, the effects and their 95% confidence intervals
  of the model are shown, where the effects are calculated by taking the exponential
  of the estimated coefficients, and all the effects are statistically significant at
  $\alpha$ level of 5%. 
  
  
  For all other incidents, the estimated delayed time is shorter compared to when
  there is a diversion, which is consistent with the patterns shown in Fig.1. For
  example, when holding other variables constant, if the delayed incident is General
  Delay, the delayed time is expected to be 79% (1 - 0.21) shorter compared to the
  delayed incident of Diversion. When holding other variables constant, if the delay
  happens on weekends (weekend = 1), the delayed time is expected to be 11% (1.11 - 1)
  longer compared to weekdays, which is consistent with the patterns shown in Fig.2. 
  Therefore, the delay-causing incident and the day of the delay both have impact 
  on the streetbus delayed minutes. 
  
  

```{r, warning = FALSE, message = FALSE, echo = FALSE}
library(lme4)
library(sjPlot)
nb_route <- glmer.nb(Min.Delay ~ Incident + weekend + (1 | Route), data = delay_final)
```

```{r, warning = FALSE, message = FALSE, echo = FALSE, fig.height = 4, fig.width = 7}
plot_model(nb_route, show.values=TRUE, show.p=TRUE, 
           title = "Fig.4 Model Effects: glmer.nb(Min.Delay ~ Incident + weekend + (1 | Route))")
```

# Discussion


  Based on the EDA and the results of the multilevel negative binomial model, there
  is strong evidence that the delay-causing incident and the delay-happening day
  both have influence on how long the delay will be. More specifically, when the 
  delay is due to diversion, the average delayed time in minutes is expected to 
  be the longest, while when the delay is due to late leaving garage, the average
  delayed time in minutes is expected to be the shortest. Also, when the delay happens
  on weekdays, the average delayed time in minutes is expected to be shorter than
  on weekends. Since this study performs a multilevel model with random intercepts
  for all 10 routes, we can also know the difference between the routes. With this
  conclusion, it might be helpful for passengers to have a general idea about how
  long they need to wait for the following streetcar.
  

### Future Directions

* Although the current multilevel negative binomial model is the most appropriate 
one after exploring other options, the residual plot and the Rootogram both suggest
that the model could be improved more. One future improvement could be doing research
and setting a delayed time cut-off point, which indicates whether the delayed time
is long or short, and then try to fit the multilevel logistic model. 

* The other improvement could be done for the future is to use the Gaussian approach
and fit the model with stan functions, which allows posterior predictive checks.

* The original data set does not contain lots of information. For future studies, 
one improvement could be adding in other potential confounding variables, such as 
the weather.  



# Appendix

### EDA 

```{r, warning = FALSE, message = FALSE, echo = FALSE, fig.cap = "The distribution of the delayed time (Min.Delay) is not very normally distributed after log transformation, therefore suggesting that a linear model is not appropriate for this data."}
hist(log(delay_final$Min.Delay + 1), main = "Distribution of The Delayed Time (min)",
     xlab = "log(Min.Delay + 1)")
```



```{r, message = FALSE, warning = FALSE, echo = FALSE}
# Difference between each delayed incident
incident <- delay %>% 
  group_by(Incident) %>% 
  filter(is.na(Min.Delay) == F) %>% 
  summarize(`mean(min)` = mean(Min.Delay), 
            `mean(hr)` = mean(Min.Delay) / 60,
            `max(min)` = max(Min.Delay),
            `max(hr)` = max(Min.Delay) / 60)
kable(incident, "simple", digits = 2, caption = "The average delayed length
      and the maximum delayed length both vary a lot among different incidents,
      which suggests that incident is a important variable to consider.")
```

```{r, warning = FALSE, message = FALSE, echo = FALSE, fig.height = 3, fig.width = 5, fig.cap = "From this plot, there is no obvious difference between the 12 months, therefore the month is not used for grouping in the model."}
month <- delay_final %>% 
  filter(Min.Delay < 60) %>% 
  filter(is.na(month) == F)

ggplot(month) +
  geom_density(alpha = 0.3) +
  aes(x = Min.Delay, color = factor(month)) + 
  facet_wrap(~ factor(month)) + 
  geom_rug() +
  xlab("Delay in Minutes") +
  theme(legend.position="none") +
  ggtitle("Distribution of the delay length among different months") +
  geom_vline(xintercept = mean(month$Min.Delay), color = "red",lty = 2) +
  theme(plot.title = element_text(hjust = 0.5))
```

## Model Fitting & Validation

```{r, warning = FALSE, message = FALSE, echo = FALSE, fig.cap = "Initial Linear Regression: lm(log(Min.Delay + 1) - Incident). Assumptions are violated, suggesting that linear model is not appropriate."}
model_lm <- lm(log(Min.Delay + 1) ~ Incident, data = delay_final)
par(mfrow=c(2,2))
 
plot(model_lm, which = 2)
plot(model_lm, which = 1) 
``` 

```{r,  echo = FALSE}
poisson <- glm(Min.Delay ~ Incident + weekend, family = "poisson", data = delay_final)
```

```{r, warning = FALSE, message = FALSE, echo = FALSE, fig.cap = "In the initial poisson regression: glm(Min.Delay - Incident + weekend, family = poisson), the model’s standard deviation is expected to have a square root relationship with the fitted mean, which can be compared with the curve in the plot. Since the majority of the points fall above the line, it indicates the existence of overdispersion, and therefore poisson model may not be the best."}
plot(fitted(poisson), resid(poisson), pch = 20)
curve(sqrt(x), add = T)
```
```{r, warning = FALSE, message = FALSE, echo = FALSE, fig.cap = "The departures from expected counts are considerably large, again suggesting the poisson model is not the best."}
# install.packages("countreg", repos="http://R-Forge.R-project.org")
library(countreg)
rootogram(poisson)
```

```{r, message = FALSE, echo = FALSE}
library(MASS)
model_nb <- glm.nb(Min.Delay ~ Incident + weekend, data = delay_final)
```

```{r, warning = FALSE, message = FALSE, echo = FALSE, fig.cap = "The initial negative binomial model is glm.nb(Min.Delay - Incident + weekend). Although there are still some points that fall above the line, the overdispersion has been fixed a lot compared to the poisson model, which suggests that the negative binomial model works better."}
plot(fitted(model_nb),resid(model_nb),pch=20)
curve(sqrt(x),add=T)
```

```{r, warning = FALSE, message = FALSE, echo = FALSE, fig.cap = "Although the roorogram does not look perfect, compared to the poisson model, the departures from expected counts are much smaller with the negative binomial model."}
rootogram(model_nb)
```


```{r, message = FALSE, echo = FALSE}
kable(cbind(Estimate = coef(model_nb), confint(model_nb)), caption = "The 95% confidence interval for the coefficients of the initial negative binomial model.")
```


```{r, warning = FALSE, message = FALSE, echo = FALSE, fig.cap = "The final multilevel negative binomial regression model is glmer.nb(Min.Delay - Incident + weekend + (1 | Route)). Although there are still some overdispersion at the lower counts, the residuals overall look fine."}
plot(fitted(nb_route), resid(nb_route),pch=20)
curve(sqrt(x),add=T)
```

```{r, echo=TRUE}
# Model Summary
summary(nb_route)
```

```{r,echo=TRUE}
# Random effects
ranef(nb_route)
```


# Bibliography

## Online Resources:

1. Wikipedia. (2020). *Toronto streetcar system* [online]. Available from: https://en.wikipedia.org/wiki/Toronto_streetcar_system#Route_numbers [accessed 29 November 2020]

2. Open Data Portal. (2020). *TTC Streetcar Delay Data* [online]. Available from: https://open.toronto.ca/dataset/ttc-streetcar-delay-data/ [accessed 5 November 2020]



## R package:

1. Garrett Grolemund, Hadley Wickham (2011). Dates and Times Made Easy with lubridate. Journal of Statistical
  Software, 40(3), 1-25. URL http://www.jstatsoft.org/v40/i03/. 
  
2. Wickham et al., (2019). Welcome to the tidyverse. Journal of Open Source Software, 4(43), 1686,
  https://doi.org/10.21105/joss.01686

3. Hadley Wickham, Romain François, Lionel Henry and Kirill Müller (2020). dplyr: A Grammar of Data
  Manipulation. R package version 1.0.2. https://CRAN.R-project.org/package=dplyr
  
4. Douglas Bates, Martin Maechler, Ben Bolker, Steve Walker (2015). Fitting Linear Mixed-Effects Models Using
  lme4. Journal of Statistical Software, 67(1), 1-48. doi:10.18637/jss.v067.i01.
  
5. Hao Zhu (2019). kableExtra: Construct Complex Table with 'kable' and
  Pipe Syntax. R package version 1.1.0.
  https://CRAN.R-project.org/package=kableExtra
  
6. Venables, W. N. & Ripley, B. D. (2002) Modern Applied Statistics with S. Fourth Edition. Springer,
  New York. ISBN 0-387-95457-0

7. Lüdecke D (2020). sjPlot: Data Visualization for Statistics in Social Science. R package version 2.8.6,
<URL: https://CRAN.R-project.org/package=sjPlot>.

8. Christian Kleiber, Achim Zeileis (2016). Visualizing Count Data Regressions Using Rootograms. The American
  Statistician, 70(3), 296--303. doi:10.1080/00031305.2016.1173590







